Model Recognition as Planning

### 2018-09-17 doc/icaps-19/model-recognition.tex
reformulating the space of possible models paper ICAPS19

  (:action dec-x
    :parameters (?x1 ?px2)
    :precondition (and (x-coord ?x1) (next ?x2 ?x1) (odd))
    :effect (and (not (x-coord ?x1)) (x-coord ?x2)))

  (:action inc-y-min-odd
    :parameters (?y1 ?y2)
    :precondition (and (y-coord ?y1) (not (next ?y2 ?y1)) (odd))
    :effect (and (not (y-coord ?y1)) (y-coord ?y2) (not (odd)) (even)))

  (:action inc-y-min-even
    :parameters (?y1 ?y2)
    :precondition (and (y-coord ?y1) (not (next ?y2 ?y1)) (even))
    :effect (and (not (y-coord ?y1)) (y-coord ?y2) (not (even)) (odd)))

  (:action inc-y-max-odd
    :parameters (?py1 ?py2)
    :precondition (and (y-coord ?y1) (not (next ?y1 ?y2)) (odd))
    :effect (and (not (y-coord ?y1)) (y-coord ?y2) (not (odd)) (even)))

  (:action inc-y-max-even
    :parameters (?y1 ?y2)
    :precondition (and (y-coord ?y1) (not (next ?y1 ?y2)) (even))
    :effect (and (not (y-coord ?y1)) (y-coord ?y2) (not (even)) (odd)))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Action model for a robot navigation in a $n\times n$ grid.}
\label{fig:model-example}
\end{figure}

@@

This paper introduces the task of {\em model recognition}.  Given a partially observed plan execution and a set of possible planning models (share the same state variables but define different action models), {\em model recognition} is the task of identifying which model in the set has the highest probability of producing the input observation. To better illustrate {\em model recognition}, imagine a robot in a $n\times n$ grid whose navigation is determined by the \strips\ model shown in Figure~\ref{fig:model-example}. According to this model the robot increments its {\em x-coordinate} when it is at an {\em even} row while, at {\em odd rows}, decrements the {\em x-coordinate}. Further, the robot exclusively increments the {\em y-coordinate} either at the {\em end} or at the {\em begining} of a row. Apart from this particular navigation model, different models can be defined within the same state variables and these models can determine different kinds of robot navigation. Given an observation of a plan execution (e.g, the six-step plan illustrated by Figure~\ref{fig:grid-example}) the {\em model recognition} task aims here to identify which navigation model produced that observation.

@@

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the dynamics of the actions $A$ (i.e. the $App$ and $\theta$ functions) are specified with \strips\ action schemas like the ones shown in Figure~\ref{fig:model-example}.

\subsection{The space of \strips\ models}
We assume that $\eff^-(a)\subseteq \pre(a)$, $\eff^-(a)\cap \eff^+(a)=\emptyset$ and $\pre(a)\cap \eff^+(a)=\emptyset$ and that actions $a\in A$ are instantiated from given action schemas, as in PDDL.

Like in PDDL~\cite{mcdermott1998pddl,fox2003pddl2}, we assume that fluents $F$ are instantiated from a set of {\em predicates} $\Psi$. Each predicate $p\in\Psi$ has an argument list of arity $ar(p)$. Given a set of {\em objects} $\Omega$, the set of fluents $F$ is induced by assigning objects in $\Omega$ to the arguments of predicates in $\Psi$, i.e.~$F=\{p(\omega):p\in\Psi,\omega\in\Omega^{ar(p)}\}$ s.t. $\Omega^k$ is the $k$-th Cartesian power of $\Omega$.

Let $\Omega_v=\{v_i\}_{i=1}^{\operatorname*{max}_{a\in A} ar(a)}$ be an additional set of objects ($\Omega\cap\Omega_v=\emptyset$), that we denote as {\em variable names}, and that is bound by the maximum arity of an action in a given planning frame. For instance, in a $5\times 5$ grid $\Omega=\{p_1, p_2, p_3, p_4, p_5\}$ while $\Omega_v=\{v_1, v_2\}$ because all the actions have arity two. We define $F_v$, a new set of fluents, $F\cap F_v=\emptyset$, produced instantiating $\Psi$ using only {\em variable names}. This set contains ten elements for the mentioned navigation model, $F_v$={\small\tt\{x-pos($v_1$), x-pos($v_2$), y-pos($v_1$), y-pos($v_2$), odd(), even(), next($v_1$,$v_1$), next($v_1$,$v_2$), next($v_2$,$v_1$), next($v_2$,$v_2$))\}}.

For a given action schema $\xi$, then $F_{\xi}\subseteq F_v$ defines the subset of elements that can appear in the preconditions and effects of that action schema. For instance $F_{\tt inc-x}=F_{\tt dec-x}=F_v$ because all actions have arity two.

Now we are ready to define the fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_{\xi}$. These fluents represent a propositional encoding for the preconditions, negative and positive effects of an action schema $\xi$ and hence, they specify the $App$ and $\theta$ functions for all the actions $a\in A$ built from the corresponding \strips\ action schemas. In more detail, if a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative/positive effect in that schema. For instance, Figure~\ref{fig:encoding} shows the conjunction of fluents representing the propositional encoding for the preconditions, negative and positive effects of the {\em inc-x} schema previously defined in Figure~\ref{fig:model-example}.

\begin{figure}
\begin{scriptsize}
\begin{verbatim}
(pre_x-pos_inc-x_v1) (pre_next_inc-x_v1_v2) (pre_even_inc-x_)
(del_x-pos_inc-x_v1) (add_x-pos_inc-x_v2)
\end{verbatim}
\end{scriptsize}
 \caption{\small Propositional encoding for the {\tt inc-x} schema previously defined in Figure~\ref{fig:model-example}.}
\label{fig:encoding}
\end{figure}

The size of the space of possible \strips\ models for an action schema $\xi$ is given by the expression, $2^{2\times|F_{\xi}|}$ (\strips\ constraints require negative effects appearing as preconditions, negative effects cannot be positive effects at the same time and also, positive effects cannot appear as preconditions). For the mentioned navigation model, $2^{2\times|F_{\xi}|}=1,048,576$ for every action schema. We say that two \strips\ schemes $\xi$ and $\xi'$ are {\em comparable} iff both share the same space of possible \strips\ models (same parameters and same constraints). For instance, we claim that all the schemas in the mentioned navigation model are {\em comparable} while {\small\tt stack} and {\small\tt pickup} from a four opertator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schema $\xi\in\mathcal{M}$ to a comparable schema $\xi'\in\mathcal{M'}$ and vice versa.

>>>>

  (:action dec-x
    :parameters (?x1 ?x2)
    :precondition (and (x-coord ?x1) (next ?x2 ?x1) (odd))
    :effect (and (not (x-coord ?x1)) (x-coord ?x2)))

  (:action inc-y-odd
    :parameters (?y1 ?y2)
    :precondition (and (y-coord ?y1) (next ?y1 ?y2) (odd))
    :effect (and (not (y-coord ?y1)) (y-coord ?y2) (not (odd)) (even)))

  (:action inc-y-even
    :parameters (?y1 ?y2)
    :precondition (and (y-coord ?y1) (next ?y1 ?y2) (even))
    :effect (and (not (y-coord ?y1)) (y-coord ?y2) (not (even)) (odd)))

  (:action dec-y-odd
    :parameters (?y1 ?y2)
    :precondition (and (y-coord ?y1) (next ?y2 ?y1) (odd))
    :effect (and (not (y-coord ?y1)) (y-coord ?y2) (not (odd)) (even)))

  (:action dec-y-even
    :parameters (?y1 ?y2)
    :precondition (and (y-coord ?y1) (next ?y2 ?y1) (even))
    :effect (and (not (y-coord ?y1)) (y-coord ?y2) (not (even)) (odd)))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Action model for a robot navigation in a $n\times n$ grid.}
\label{fig:model-example}
\end{figure}

@@

This paper introduces the task of {\em model recognition}.  Given a partially observed plan execution and a set of possible planning models (share the same state variables but define different action models), {\em model recognition} is the task of identifying which model in the set has the highest probability of producing the input observation. To better illustrate {\em model recognition}, imagine a robot in a $n\times n$ grid whose navigation is determined by the \strips\ model shown in Figure~\ref{fig:model-example}. According to this model the robot can increment its {\em x-coordinate} when it is at an {\em even} row while, at {\em odd rows}, can decrement the {\em x-coordinate}. Apart from this particular navigation model, different models can be defined within the same state variables and these models can determine different kinds of robot navigation. Given an observation of a plan execution (e.g, the six-step plan illustrated by Figure~\ref{fig:grid-example}) the {\em model recognition} task aims here to identify which navigation model produced that observation.

@@

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the dynamics of the actions $A$ (i.e. the $App$ and $\theta$ functions) are specified with \strips\ action schemas. We start formalizing the full space of possible \strips\ models, next we formalize an {\em edit distance} for \strips\ models and eventually, we define the $P(\mathcal{M}|\tau)$ probability distribution for \strips\ models.

\subsection{The space of \strips\ models}
An \strips\ schema $\xi$ is defined by three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the fluents that can appear in the preconditions, negative effects and positivie effects of an /strips/ action.

We say that two \strips\ schemes $\xi$ and $\xi'$ are {\em comparable} iff both share the same parameters and same constraints. For instance, we claim that all the schemas in the mentioned navigation model are {\em comparable} while {\small\tt stack} and {\small\tt pickup} from a four opertator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schema $\xi\in\mathcal{M}$ to a comparable schema $\xi'\in\mathcal{M'}$ and vice versa.

The space of possible \strips\ models is bound by these three elements:
\begin{itemize}
\item $\Psi$, is the set of {\em state predicates}. A set of FOL predicates  that shape the propositional state varibles.
\item $\Psi\cap\Psi_a =\emptyset$, is a set of additional predicates that shape the action headers. An action header of an action schema $\xi$ is defined by a predicate $p_\xi=name_\xi(pars_\xi)$ such that $name_\xi$ is the name of action schema $\xi$ and $pars$ is the list of the corresponding parameters. For instance, the set of predictes induced from the PDDL schema shown in Figure~\ref{fig:model-example} is $\{${\small\tt inc-x(X1,X2), dec-x(X1,X2), inc-y-odd(X1,X2), inc-y-even(X1,X2), inc-y-odd (X1,X2), inc-y-even(X1,X2)}$\}$. 
\item ${\mathcal C}$, a set of {\em syntactic constraints}. These constraints include the \strips\ constraints (we assume that $\eff^-(a)\subseteq \pre(a)$, $\eff^-(a)\cap \eff^+(a)=\emptyset$ and $\pre(a)\cap \eff^+(a)=\emptyset$) and domain-specific constraints. For instace, in a {\em robot navigation} domain like the modeled in Figure~\ref{fig:model-example}, predicates {\small\tt even()} and {\small\tt odd()} are exclusive so they cannot hold a the same time.
\end{itemize}

Predicates $\Psi_a$ also shape an additional set of objects ($\Omega\cap\Omega_v=\emptyset$) that we denote as {\em variable names} and that is bound by the maximum arity of the given action headers, i.e., $\Omega_v=\{v_i\}_{i=1}^{\operatorname*{max}_{\xi} ar(\xi)}$. For instance, in a $5\times 5$ grid (like the one in Figure~\label{fig:grid-example} and modeled in Figure~\ref{fig:model-example}) $\Omega=\{o_1, o_2, o_3, o_4, o_5\}$ while $\Omega_v=\{v_1, v_2\}$ because all the actions schema have arity two.

Given the set of {\em state predicates} $\Psi$ and the set of {\em variable names} $\Omega_v$, then $F_{\xi}\subseteq F_v$ defines the subset of elements that can appear in the preconditions and effects of that action schema $\xi$. This set is the same contains ten elements for the mentioned navigation model, $F_v$={\small\tt\{x-coord($v_1$), x-coord($v_2$), y-coord($v_1$), y-coord($v_2$), odd(), even(), next($v_1$,$v_1$), next($v_1$,$v_2$), next($v_2$,$v_1$), next($v_2$,$v_2$))\}}.

\begin{definition}[The space of \strips\ models]
Given the set of {\em state predicates} $\Psi$, the set of {\em variable names} $\Omega_v$ and the set of {\em syntactic constraints} ${\mathcal C}$, the space of models for a given op predicate $p_\xi$ that represents the header of an action schema $\xi$ is given by three sets of inter
\end{definition}

The size of the space of possible \strips\ models for an action schema $\xi$ is given by the expression, $2^{2\times|F_{\xi}|}$ (\strips\ constraints require negative effects appearing as preconditions, negative effects cannot be positive effects at the same time and also, positive effects cannot appear as preconditions). For the mentioned navigation model, $2^{2\times|F_{\xi}|}=1,048,576$ for every action schema.

\subsection{A propositional encoding for \strips\ models}
Now we are ready to define the fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_{\xi}$. These fluents represent a propositional encoding for the preconditions, negative and positive effects of an action schema $\xi$ and hence, they specify the $App$ and $\theta$ functions for all the actions $a\in A$ built from the corresponding \strips\ action schemas. In more detail, if a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative/positive effect in that schema. For instance, Figure~\ref{fig:encoding} shows the conjunction of fluents representing the propositional encoding for the preconditions, negative and positive effects of the {\em inc-x} schema previously defined in Figure~\ref{fig:model-example}.

\begin{figure}
\begin{tiny}
\begin{verbatim}
;;; Propositional encoding for inc-x(?x1 ?p2)
(pre_x-coord_inc-x_v1) (pre_next_inc-x_v1_v2) (pre_even_inc-x_)
(del_x-coord_inc-x_v1) (add_x-coord_inc-x_v2)

;;; Propositional encoding for dec-x(?x1 ?p2)
(pre_x-coord_inc-x_v1) (pre_next_inc-x_v2_v1) (pre_odd_inc-x_)
(del_x-coord_inc-x_v1) (add_x-coord_inc-x_v2)

;;; Propositional encoding for inc-y-odd(?y1 ?y2)
(pre_y-coord_inc-y-odd_v1) (pre_next_inc-y-odd_v1_v2) 
(pre_even_inc-y-odd_)
(del_y-coord_inc-y-odd_v1) (del_odd_inc-y-odd_)
(add_y-coord_inc-y-odd_v2) (add_even_inc-y-odd_)

;;; Propositional encoding for inc-y-even(?y1 ?y2)
(pre_y-coord_inc-y-even_v1) (pre_next_inc-y-even_v1_v2)
(pre_even_inc-y-even_)
(del_y-coord_inc-y-even_v1) (del_even_inc-y-even_)
(add_y-coord_inc-y-even_v2) (add_odd_inc-y-even_)

;;; Propositional encoding for dec-y-odd(?y1 ?y2)
(pre_y-coord_dec-y-odd_v1) (pre_next_dec-y-odd_v2_v1)
(pre_even_dec-y-odd_)
(del_y-coord_dec-y-odd_v1) (del_odd_dec-y-odd_)
(add_y-coord_dec-y-odd_v2) (add_even_dec-y-odd_)

;;; Propositional encoding for inc-y-even(?y1 ?y2)
(pre_y-coord_dec-y-even_v1) (pre_next_dec-y-even_v2_v1)
(pre_even_dec-y-even_)
(del_y-coord_dec-y-even_v1) (del_even_dec-y-even_)
(add_y-coord_dec-y-even_v2) (add_odd_dec-y-even_)

\end{verbatim}
\end{tiny}
 \caption{\small Propositional encoding for the action schema previously defined in Figure~\ref{fig:model-example}.}
\label{fig:encoding}
\end{figure}
